Detecting Interactions in Random Forest for Genomic Data

### 2025-04-30 manuscript.tex
changes for elsarticle style

@@

In addition it is clear that Random Forests can perform well with correlated and interacting variables, at least in the
sense that the fitted model performs well \cite[]{Ziegler.et.al.2007}. However, most of the Random Forest based
approaches are either based on single variable importance \cite[]{Jiang.et.al.2009, yoshidaSNPInterForestNewMethod2011}
or based on pairwise variable importance \cite[]{Ziegler.et.al.2007}. The Random Forest-based approaches for identifying
interactions based on variable importance measurements cannot distinguish whether a discovered interaction is a true
interaction or simply two variables with strong marginal effects. Additionally, if the interacting variables have small
marginal effects, they will not appear near the top of the variable importance list. This means that detecting
interactions based on variable importance can be problematic. It is also heavily dependent on extensive permutation
tests.  \citeauthor{schmalohrDetectionEpistaticInteractions2018} notes that there is a lack of accurate methods for the
detection of interactions between genomic loci that affect QTLs but that several studies have shown that RF outperforms
other QTL mapping methods at identifying QTL and discrete trait associations. They provided benchmarks on simulated and
real yeast cross data indicating that the RF-based methods outperform other commonly used approaches for detecting
epistasis.

@@

@@

Working from the random forest output, we can form a $p \times p$ matrix of values such that the $\{i,j\}^{th}$ entry is the
count of the number of times variable $v_i$ is followed by variable $v_j$.  If we sum the upper and lower triangular matrices
and set the diagonal to 0, we form the interaction vector, of length $\frac {p(p-1)} 2 = {p \choose 2}$.
This is likely to be large ($4 \times 10^9$ in the rheumatism example) and contain a large number of zeros (0.999 zeros
in the rheumatism example).

Efron's Empirical Bayes procedure (\cite{efronLargeScaleInferenceEmpirical2010}) is now widely used to determine levels
of significance in high throughput genomics data analysis.
These data sets led to the use of large numbers of parallel tests to determine the significant variables and then
the use of multiple testing corrections \cite{limma.2015}.
Efron suggests that instead of assuming a distribution, such as the $t$, 
a the better approach is to use the large number of variables to model the null distribution, and then do
significance testing.

@@

We model the data as 
\[
f (j ) = \pi_0 f_0 (j ) + (1 - \pi_0 )f_1 (j )
\]
where $j$ is the count data, $f_0$ is the null distribution and $f_1$ is some contaminating distribution of the
non-null counts. Following \cite{Gauran.et.al.2018} we could make the additional assumption that there is some integer $C$ such that
\[
  f(j ) = \pi_0 f_0 (j ) \textrm{ for } j \leq C.
\]
So we only have the null distribution for counts less than $C.$

As we have count data we model the null distribution $f_0$ as a Poisson or  generalized Poisson ($g$, with parameters
$\lambda$ and $\theta$).
If zero is observed with a significantly higher frequency, as here, we can include a zero-inflation parameter in the
distribution, then $J \sim ZIGP(\eta, \lambda, \theta)$ (zero-inflated generalized Poisson) and the probability that $J = j$, denoted by $f_0 (j)$, 
\[
  f_0(j) = \eta I_{\{0\}} (j) + (1 - \eta) g(j) I_{\{0,1,2,\ldots \}} (j)
\]
where $j$ is a non-negative integer, $0 \leq \eta < 1$, $0 \leq \theta < 1$ and $\lambda > 0$. The indicator function
$I_S (j)$ is equal to 1 if $j \in  S$ and 0 otherwise. Fitting this model involves estimating $\{ \eta, \pi_0,
\lambda$ and $\theta \}$. Having estimated the parameters we choose a cutoff value $D_n$, chosen to control the false
discovery rate, so that counts $\geq  D_n$ are reported as significant.

@@

There are several options described in  \cite{Gauran.et.al.2018}.
Using the ``one stage procedure'' and $C_1$ constant (see supplementary information)
we get a $C$ value (ref equation) of 6 and and a $D_n$ value of 7 which gives us 57 significant interaction.

We note an important caveat here.  \cite{Gauran.et.al.2018} show that the significance level can be
determined and the FDR controlled in a series of simulation examples. For example  Figure 1 (\cite{Gauran.et.al.2018})
shows an example where $\lambda =20$ for the null distribution and the contaminating distribution is a Binomial with
$\mu=50.$
It would appear that these examples all have $\lambda$ and $\mu$ (the mean of the contaminating distribution) with a
greater separation than we have in the Rheumatoid data.
Now, somewhat uniquely, we can move the contaminating distribution. All we need to do is fit a Random Forest with more
trees and we will observe counts.

It appears that there is a trade-off  here. And as we are sampling a space of $4 \times 10^{9}$ possible interactions,
we might have to increase the number of trees a lot to get a meaningful difference in the mean of the contaminating
distribution.
In this case we have followed a simple strategy:
\begin{itemize}
\item threshold the observed count data. We do this a number of ways to do this (list them) and then take the median. IN
  the case of the Rheumatoid data, this was 2 which gave us 222 interactions;
\item test each interaction with a linear model (see below).
\end{itemize}

@@

\section{Methods}
% The proposed method is described in Figure \ref{figure:RF_Method}. First, a Random Forest model was created using all avaiable data. Then for each pair 
% of SNP, how many times they appeared as parent-child node in each of the decision trees were recorded. Then the SNP pairs were ranked based on
% the count and only the pairs having count >= 2 were retained. Finally, using the linier equation the interaction term was tested. In our
% studies, we used the ranger R package \cite[]{Wright.and.Ziegler.2017} and we selected \texttt{num.tree} = 1000 and \texttt{mtry} =
% \#SNPs/3.
% For a linear model, an interaction occurs when an independent variable has a different effect on the outcome depending
% on the values of another independent variable. So our variables in the model can be divided between main effects and
% interaction effects. A multiplicative epistasis model was used, where the interaction between two SNPs, $i$ and $j$,  were modelled as the
% product of the two genotype values, using the following linear regression model:

The proposed method is described in Figure \ref{figure:RF_Method}. First, a Random Forests model is created using all
available genomic data. For both the simulated and real data examples, the R package ranger
\cite[]{Wright.and.Ziegler.2017} was used with the parameters \texttt{num.tree} = 1000 and \texttt{mtry} = \#SNPs/3.
Once the Random Forests model is built, the number of times each pair of markers was found as parent-child nodes in all
the decision trees were recorded. Then the marker pairs were ranked based on the count and only the pairs having count
more than 2 were retained. A threshold of at least two ($t = 2$) appearances as parent-child nodes in the decision trees
of the Random Forests was used because the use of one ($t = 1$) resulted in a substantially higher number of SNP pairs,
often random occurrences, that did not improve the identification of interactions.  Finally, using linear models the
significance of the interactions are tested. For a linear model, an interaction occurs when an independent variable has
a different effect on the outcome depending on the values of another independent variable. The variables in the model
are divided between main and interaction effects. A multiplicative epistasis model was used, where the interaction
between two markers, $i$ and $j$, are modelled as the product of the two genotype values, using the following linear
regression model:

@@

\subsection{Real datasets -- Wellcome Trust Case Control Consortium data for rheumatoid arthritis}
The first dataset was obtained from the Wellcome Trust Case Control Consortium (WTCCC)
\cite[]{burtonGenomewideAssociationStudy2007brief}. The original dataset contains 14,000 cases for seven complex human
diseases and 3,000 shared controls. The control individuals were from two different sources: 1958 British Birth Cohort
(58C) and the UK Blood Service Control Group (UKBS). Each of the control group has 1,500 individuals. Samples of DNA
from all 17,000 individuals were genotyped with the GeneChip 500K Mapping Array Set (Affymetrix chip).  The current
analysis only used the data for rheumatoid arthritis (RA) that consisted of 4,798 individuals with 1,860 cases and 2,938
controls. Variants having $>10\%$ missing values and MAFs$<0.05$ were removed. SNP markers that failed the
Hardy-Weinberg test (p $\leq$ 0.05) were also removed. As linkage disequilibrium (LD) could affect how the Random Forest
(RF) algorithm selects a SNP at a particular node (for more detail how LD can affect RF, please see
\cite{mengPerformanceRandomForest2009}), the dataset was also filtered based on LD. Using PLINK 1.9
\cite[]{purcellPLINKToolSet2007} and a sliding window of 200 kb, one SNP from a pair of SNPs in that window was removed
if pairwise $r^2 >=0.3$.

>>>>

\tableofcontents 
\clearpage

@@

In addition it is clear that Random Forests can perform well with correlated and interacting variables, at least in the
sense that the fitted model performs well \cite[]{Ziegler.et.al.2007}. However, most of the Random Forest based
approaches are either based on single variable importance \cite[]{Jiang.et.al.2009, yoshidaSNPInterForestNewMethod2011}
or based on pairwise variable importance \cite[]{Ziegler.et.al.2007}. The Random Forest-based approaches for identifying
interactions based on variable importance measurements cannot distinguish whether a discovered interaction is a true
interaction or simply two variables with strong marginal effects. Additionally, if the interacting variables have small
marginal effects, they will not appear near the top of the variable importance list. This means that detecting
interactions based on variable importance can be problematic. It is also heavily dependent on extensive permutation
tests.  \cite{schmalohrDetectionEpistaticInteractions2018} notes that there is a lack of accurate methods for the
detection of interactions between genomic loci that affect QTLs but that several studies have shown that RF outperforms
other QTL mapping methods at identifying QTL and discrete trait associations. They provided benchmarks on simulated and
real yeast cross data indicating that the RF-based methods outperform other commonly used approaches for detecting
epistasis.

@@

\section{Methods}

The proposed method is described in Figure \ref{figure:RF_Method}. First, a Random Forests model is created using all
available genomic data. For both the simulated and real data examples, the R package ranger
\cite[]{Wright.and.Ziegler.2017} was used with the parameters \texttt{num.tree} = 1000 and \texttt{mtry} = \#SNPs/3.
%Once the Random Forests model is built, the number of times each pair of markers was found as parent-child nodes in all
%the decision trees were recorded,
Working from the random forest output, we can form a $p \times p$ matrix of values such that the $\{i,j\}^{th}$ entry is the count of the
number of times variable $v_i$ is followed by variable $v_j$.  If we sum the upper and lower triangular matrices and set the diagonal to 0,
we form the interaction vector, of length $\frac {p(p-1)} 2 = {p \choose 2}$.  This is likely to be large and contain a large number of
zeros ($4 \times 10^9$ entries and 0.999 zero in the rheumatoid example, section \ref{subsection: rheumatoid}).

@@

Efron's Empirical Bayes procedure \cite[]{efronLargeScaleInferenceEmpirical2010} is now widely used to determine levels of significance in
high throughput genomics data analysis.  These data sets led to the use of large numbers of parallel tests to determine the significant
variables and the use of multiple testing corrections \cite{limma.2015}.  Efron suggests that instead of assuming a distribution, such as
the $t$, the better approach is to use the large number of variables to model the null distribution, and then do significance testing.

@@

We model the data as 
\[
f (j ) = \pi_0 f_0 (j ) + (1 - \pi_0 )f_1 (j )
\]
where $j$ is the count data, $f_0$ is the null distribution and $f_1$ is some contaminating distribution of the
non-null counts. Following \cite{Gauran.et.al.2018} we could make the additional assumption that there is some integer $C$ such that
\begin{equation}
  f(j ) = \pi_0 f_0 (j ) \textrm{ for } j \leq C.
\label{eqn:C}  
\end{equation}
So we only have the null distribution for counts less than $C.$

As we have count data we model the null distribution $f_0$ as a Poisson or  generalized Poisson ($g$, with parameters
$\lambda$ and $\theta$).
If zero is observed with a significantly higher frequency, as here, we can include a zero-inflation parameter in the
distribution, then $J \sim ZIGP(\eta, \lambda, \theta)$ (zero-inflated generalized Poisson) and the probability that $J = j$, denoted by $f_0 (j)$, 
\[
  f_0(j) = \eta I_{\{0\}} (j) + (1 - \eta) g(j) I_{\{0,1,2,\ldots \}} (j)
\]
where $0 \leq \eta < 1$, $0 \leq \theta < 1$ and $\lambda > 0$. The indicator function
$I_S (j)$ is equal to 1 if $j \in  S$ and 0 otherwise. Fitting this model involves estimating $\{ \eta, \pi_0,
\lambda$ and $\theta \}$. Having estimated the parameters we choose a cutoff value $D_n$, chosen to control the false
discovery rate, so that counts $\geq  D_n$ are reported as significant.

@@

There are several options described in  \cite{Gauran.et.al.2018}.
Using the ``one stage procedure'' and $C_1$ constant (see supplementary information)
we get a $C$ value (equation \ref{eqn:C}) of 6 and and a $D_n$ value of 7 which gives us 57 significant interaction.

We note an important caveat here.  \cite{Gauran.et.al.2018} show that the significance level can be
determined and the FDR controlled in a series of simulation examples. For example  Figure 1 \cite[]{Gauran.et.al.2018}
shows an example where $\lambda =20$ for the null distribution and the contaminating distribution is a Binomial with
$\mu=50.$ It would appear that these examples all have $\lambda$ and $\mu$ (the mean of the contaminating distribution) with a
greater separation than we have in the Rheumatoid data. So we may have a more inteactable problem here.

Now, somewhat uniquely, we can move the contaminating distribution. All we need to do is fit a Random Forest with more
trees and we will observe more counts.
There is a trade-off here and as we are sampling a space with typically more than $10^{9}$ possible interactions,
we have to increase the number of trees a lot to get a meaningful difference in the mean of the contaminating
distribution. Because of this we resort to a threshholding stratey that is both computationally easier and controls the FDR rate by testing
each variable.

\subsection{threshholding the data}
In this case we have followed a simple strategy:
\begin{itemize}
\item threshold the observed count data. We do this a number of ways to do this (list them) and then take the median. In
  the case of the Rheumatoid data, this was 2 which gave us 222 interactions;
\item test each interaction with a linear model (see below).
\end{itemize}

@@

% The proposed method is described in Figure \ref{figure:RF_Method}. First, a Random Forest model was created using all avaiable data. Then for each pair 
% of SNP, how many times they appeared as parent-child node in each of the decision trees were recorded. Then the SNP pairs were ranked based on
% the count and only the pairs having count >= 2 were retained. Finally, using the linier equation the interaction term was tested. In our
% studies, we used the ranger R package \cite[]{Wright.and.Ziegler.2017} and we selected \texttt{num.tree} = 1000 and \texttt{mtry} =
% \#SNPs/3.
% For a linear model, an interaction occurs when an independent variable has a different effect on the outcome depending
% on the values of another independent variable. So our variables in the model can be divided between main effects and
% interaction effects. A multiplicative epistasis model was used, where the interaction between two SNPs, $i$ and $j$,  were modelled as the
% product of the two genotype values, using the following linear regression model:

Then the marker pairs were ranked based on the count and only the pairs having count
more than 2 were retained. A threshold of at least two ($t = 2$) appearances as parent-child nodes in the decision trees
of the Random Forests was used because the use of one ($t = 1$) resulted in a substantially higher number of SNP pairs,
often random occurrences, that did not improve the identification of interactions.  Finally, using linear models the
significance of the interactions are tested. For a linear model, an interaction occurs when an independent variable has
a different effect on the outcome depending on the values of another independent variable. The variables in the model
are divided between main and interaction effects. A multiplicative epistasis model was used, where the interaction
between two markers, $i$ and $j$, are modelled as the product of the two genotype values, using the following linear
regression model:

@@

\subsection{Real datasets -- Wellcome Trust Case Control Consortium data for rheumatoid arthritis}
\label{subsection: rheumatoid}
The first dataset was obtained from the Wellcome Trust Case Control Consortium (WTCCC)
\cite[]{burtonGenomewideAssociationStudy2007brief}. The original dataset contains 14,000 cases for seven complex human
diseases and 3,000 shared controls. The control individuals were from two different sources: 1958 British Birth Cohort
(58C) and the UK Blood Service Control Group (UKBS). Each of the control group has 1,500 individuals. Samples of DNA
from all 17,000 individuals were genotyped with the GeneChip 500K Mapping Array Set (Affymetrix chip).  The current
analysis only used the data for rheumatoid arthritis (RA) that consisted of 4,798 individuals with 1,860 cases and 2,938
controls. Variants having $>10\%$ missing values and MAFs$<0.05$ were removed. SNP markers that failed the
Hardy-Weinberg test (p $\leq$ 0.05) were also removed. As linkage disequilibrium (LD) could affect how the Random Forest
(RF) algorithm selects a SNP at a particular node (for more detail how LD can affect RF, please see
\cite{mengPerformanceRandomForest2009}), the dataset was also filtered based on LD. Using PLINK 1.9
\cite[]{purcellPLINKToolSet2007} and a sliding window of 200 kb, one SNP from a pair of SNPs in that window was removed
if pairwise $r^2 >=0.3$.
